---
title: "clusterProfiler-specific-MSigDB"
author: "Eileen Owens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro
The purpose of this script is to perform and plot GSEA conducted with the R package clusterProfiler. This analysis evaluates the enrichment of user-supplied differential gene expression data for gene sets available in the Molecular Signatures Database, and allows the selection of specific MSigDB categories for analysis.

```{r installation, include=FALSE}
# Load necessary packages
library(clusterProfiler)
library(pathview)
library(enrichplot)
library(ggplot2)
library(ggnewscale)
library(msigdbr)
library(DOSE)
library(ReactomePA)
library(tidyverse)
library(GSEABase)
library(RColorBrewer)
```


```{r data-input, echo=FALSE}
# Set working directory:
setwd("~/PTCL/PROJ01_26PTCLs/02_scripts")
# Read in data from deseq2
df = read.csv("../01_input/CD4PTCLvsCD4CTRL_DESeq2res_BlankIDsRemoved.csv", header=TRUE)

```

```{r convert-probeid, eval=FALSE, echo=FALSE}
# If there is no column of gene symbols (probe ID only), run this code. Otherwise, exclude this code chunk with eval=FALSE and run the next code chunk instead.

# take the stat column from DESeq2 results for ranking
original_gene_list <- df$stat
# name the vector of probe IDs (ENSEMBL gene IDs)
names(original_gene_list) <- df$X
# omit any NA values
gene_list <- na.omit(original_gene_list)
# sort the list in decreasing order (required for clusterProfiler)
gene_list = sort(gene_list, decreasing = TRUE)

# Install annotations
## Install and load the canine genome annotations. All available annotations can be found here: http://bioconductor.org/packages/release/BiocViews.html#___OrgDb.
organism = "org.Cf.eg.db" # Set the desired organism.
# BiocManager::install(organism, character.only=TRUE) # First time only, then comment out this line.
library(organism, character.only=TRUE)

keytypes(org.Cf.eg.db)
symbols <- bitr(names(original_gene_list), fromType="ENSEMBL", toType="SYMBOL", OrgDb=organism)
dedup_symbols <- symbols[!duplicated(symbols[c("ENSEMBL")]),] # Removes duplicate IDs
df2 <- df[df$X %in% dedup_symbols$ENSEMBL,] # Creates a new dataframe, df2, which has only the genes which were successfully mapped using the bitr function.
df2$Y <- dedup_symbols$SYMBOL # Creates a new column in df2 with the corresponding gene symbol
deg_genes_list <- df2$stat # Creates a vector of the stat column from DESeq2 output
names(deg_genes_list) <- df2$Y # Names vector with gene symbols
deg_genes_list <- na.omit(deg_genes_list) # Omits any NA values
deg_genes_list = sort(deg_genes_list, decreasing = TRUE) # Sorts in decreasing order
head(deg_genes_list)
deg_names <- names(deg_genes_list) # Makes a vector of just the ranked gene names

```

```{r}
# If there is a column of gene symbols in the imported deseq2 data, run this code chunk. Otherwise, exclude with eval=FALSE and run the preceding code chunk.

df <- df[!duplicated(df$gene_name)] # Remove rows with duplicate gene symbols.
deg_genes_list <- df$stat # Creates a vector of the stat column from DESeq2 output
names(deg_genes_list) <- df$gene_name # Names vector with gene symbols
deg_genes_list <- na.omit(deg_genes_list) # Omits any NA values
deg_genes_list = sort(deg_genes_list, decreasing = TRUE) # Sorts in decreasing order
head(deg_genes_list)
deg_names <- names(deg_genes_list) # Makes a vector of just the ranked gene names

```

```{r msigdb-import, echo=FALSE}
## Retrieve the specific collection of interest from MSigDB.
# H = hallmark gene sets, C1 = positional gene sets, C2 = curated gene sets, C3 = motif gene sets, C4 = computational gene sets, C5 = GO gene sets, C6 = oncogenic signatures, C7 = immunologic signatures.
m_t2g <- msigdbr(species = "Homo sapiens") %>%
  dplyr::select(gs_name, gene_symbol)
head(m_t2g)

```

```{r gsea, echo=FALSE}
## GSEA using MSigDb gene sets
gse_msigdb <- GSEA(deg_genes_list, # Use the sorted (ranked) list of gene symbols *and* stat ranking column
                  exponent = 1,
                  nPerm = 10000,
                  minGSSize = 1,
                  maxGSSize = 800,
                  pvalueCutoff = 0.05,
                  pAdjustMethod = "BH",
                  TERM2GENE = m_t2g,
                  verbose = TRUE,
                  by = "fgsea")

# Export results:
gse_msigdb_df <- as.data.frame(gse_msigdb)
write.csv(gse_msigdb_df, file="../03_output/<filename>.csv")

## MSigDB over-presentation analysis. For this, we need a vector of gene IDs.
# To select only those genes with log2fc > 1.5:
genes_mostup <- names(deg_genes_list)[abs(deg_genes_list) > 1.5]
# To create a vector of only the top 250 upregulated genes:
top250_genelist <- df_top250$stat
top250_geneVector <- df_top250$gene_name

enriched_msigdb <- enricher(top250_geneVector, TERM2GENE=m_t2g)

# Export results
enriched_msigdb_df <- as.data.frame(enriched_msigdb)
write.csv(enriched_msigdb_df, file="../03_output/enricherMSigDB_top250.csv")
```

## Dotplot
```{r}
dotplot(gse_msigdb, showCategory=10, split=".sign") + facet_grid(.~.sign)
```

## Ridgeplot
```{r}
ridgeplot(gse_msigdb, showCategory=20) + labs(x = "enrichment distribution")
```

# Citations
```{r}
sessionInfo()
citation()
```
