---
title: "Clustering Analysis of Canine PTCL Bulk RNA-seq Data"
author: "Eileen Owens"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      cache = TRUE,
                      keep_md = TRUE)
```

```{r load-packages, include=FALSE}
library(DESeq2)
library(RColorBrewer)
library(pheatmap)
library(dplyr)
library(readr)
library(knitr)
library(ggplot2)
```

```{r working-directory, include=FALSE}
# Set your working directory
setwd("C:/Users/edlarsen/Documents/PTCLRNASeq")
```

# Introduction
This script performs clustering and draws heatmaps of RNA-seq data, with a variety of filtering options.

# Data
### Variance stabilized transformed count matrix
```{r data-input-vst, eval=TRUE}
vsd_cohort1 <- read.csv("Cohort_1/Output/VST_NormalizedCounts.csv")
vsd_cohort2 <- read.csv("Cohort_2/Output/VST_NormalizedCounts.csv")

## Make gene names the row names
rownames(vsd_cohort1) <- make.unique(as.character(vsd_cohort1$gene_name)) # make.unique accounts for duplicate gene names since R won't allow duplicate rownames
rownames(vsd_cohort2) <- make.unique(as.character(vsd_cohort2$gene_name))

## Remove any columns other than sample count data
vsd_cohort1 <- dplyr::select(vsd_cohort1, -c("X", "probe_id", "gene_name", "description")) # exclude any columns that are not sample IDs
vsd_cohort2 <- dplyr::select(vsd_cohort2, -c("X", "probe_id", "gene_name", "description")) # exclude any columns that are not sample IDs
```

### Metadata
```{r metadata}
metadata1 <- read.table(file = "Cohort_1/Input/metadata.txt", header = FALSE)
metadata2 <- read.table(file = "Cohort_2/Input/metadata.txt", header = FALSE)
colnames(metadata1) <- c("avery_num", "sample_name", "phenotype")
colnames(metadata2) <- c("avery_num", "sample_name", "phenotype")
rownames(metadata1) <- metadata1$sample_name
metadata1 <- dplyr::select(metadata1, c("phenotype"))
rownames(metadata2) <- metadata2$sample_name
metadata2 <- dplyr::select(metadata2, c("phenotype"))
```

# Unsupervised clustering of the top 2000 genes (by median absolute derivation)
```{r mad}
# Calculate the median absolute derivation for all rows in the vst transformed data
median_absolute_derivation_cohort1 = apply(vsd_cohort1,1,mad) # "1" indicates that the manipulation is performed on rows
median_absolute_derivation_cohort2 = apply(vsd_cohort2,1,mad)

# check data distribution
hist(median_absolute_derivation_cohort1, ylim=c(0,200), main=c("Histogram of median absolute derivation, Cohort 1"), breaks=nrow(vsd_cohort1)*0.1)
hist(median_absolute_derivation_cohort2, ylim=c(0,200), main=c("Histogram of median absolute derivation, Cohort 2"), breaks=nrow(vsd_cohort2)*0.1)

# index the vst transformed count data to include only those rows (genes) that appeared in the top 2000 based on median absolute derivation.
mad2k_cohort1=vsd_cohort1[rev(order(median_absolute_derivation_cohort1))[1:2000],]
mad2k_cohort2=vsd_cohort2[rev(order(median_absolute_derivation_cohort1))[1:2000],]
```

## Euclidean distance
```{r mad-heatmap-euclidean-cohort1, fig.width=14}
# Specify colors (optional). Full list of color options here: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
ann_colors_cohort1 = list(
  phenotype = c("CD8_PTCL" = "deepskyblue", "DN_PTCL" = "magenta2", "CD8_CTRL" = "seagreen3", "CD4_PTCL" = "darkkhaki", "CD4_CTRL" = "coral"))

# Draw heatmap
mad_heatmap_cohort1 <- pheatmap(mad2k_cohort1, 
              scale="row",
              color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
              cluster_rows=TRUE,
              cluster_cols=TRUE,
              #cutree_rows = 3,
              cutree_cols = 3,
              main = "Input: Vst transformed normalized DESeq2 counts for the top 2000 genes by median absolute derivation, Cohort 1 \n Clustering: Ward, Distance: Euclidean",
              clustering_distance_rows = "euclidean",
              clustering_distance_cols = "euclidean",
              clustering_method = "ward.D2",
              annotation_col = metadata1,
              annotation_colors = ann_colors_cohort1,
              show_rownames = FALSE) # Change to TRUE if gene symbols should be annotated on the heatmap.
```

```{r mad-heatmap-euclidean-cohort2, fig.width=14}
# Specify colors (optional). Full list of color options here: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
ann_colors_cohort2 = list(
  phenotype = c("CD4_LN_CTRL" = "coral", "CD4_PTCL" = "darkkhaki", "CD4THYM_CTRL" = "deepskyblue"))

# Draw heatmap
mad_heatmap_cohort2 <- pheatmap(mad2k_cohort2, 
              scale="row",
              color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
              cluster_rows=TRUE,
              cluster_cols=TRUE,
              #cutree_rows = 3,
              cutree_cols = 3,
              main = "Input: Vst transformed normalized DESeq2 counts for the top 2000 genes by median absolute derivation, Cohort 2 \n Clustering: Ward, Distance: Euclidean",
              clustering_distance_rows = "euclidean",
              clustering_distance_cols = "euclidean",
              clustering_method = "ward.D2",
              annotation_col = metadata2,
              annotation_colors = ann_colors_cohort2,
              show_rownames = FALSE) # Change to TRUE if gene symbols should be annotated on the heatmap.
```

## Pearson correlation
```{r mad-heatmap-pearson-cohort1, fig.width=14}
mad_heatmap_cohort1 <- pheatmap(mad2k_cohort1, 
              scale="row",
              color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
              cluster_rows=TRUE,
              cluster_cols=TRUE,
              #cutree_rows = 2,
              cutree_cols = 3,
              main = "Input: Vst transformed normalized DESeq2 counts for the top 2000 genes by median absolute derivation, Cohort 1\n Clustering: Ward, Distance: Pearson correlation",
              clustering_distance_rows = "correlation",
              clustering_distance_cols = "correlation",
              clustering_method = "ward.D2",
              annotation_col = metadata1,
              annotation_colors = ann_colors_cohort1,
              show_rownames = FALSE) # Change to TRUE if gene symbols should be annotated on the heatmap.
```

```{r mad-heatmap-pearson-cohort2, fig.width=14}
mad_heatmap_cohort2 <- pheatmap(mad2k_cohort2, 
              scale="row",
              color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
              cluster_rows=TRUE,
              cluster_cols=TRUE,
              #cutree_rows = 2,
              cutree_cols = 3,
              main = "Input: Vst transformed normalized DESeq2 counts for the top 2000 genes by median absolute derivation, Cohort 2\n Clustering: Ward, Distance: Pearson correlation",
              clustering_distance_rows = "correlation",
              clustering_distance_cols = "correlation",
              clustering_method = "ward.D2",
              annotation_col = metadata2,
              annotation_colors = ann_colors_cohort2,
              show_rownames = FALSE) # Change to TRUE if gene symbols should be annotated on the heatmap.
```

# Clustering based on expression of human PTCL gene signatures
## Iqbal et al.
Source: https://doi.org/10.1182/blood-2013-11-536359
```{r geneList}
# Define list of genes and their groups
geneList <- c("TBX21", "IFNG", "STAT1", "CSF2", "CCL3", "EOMES", "IL2RB", "CXCR3",
              "GATA3", "EGR1", "SEPTIN6", "CAT", "IL18R1", "IK", "ACKR3", "CCR4", "MSH6")
geneGroups <- c("TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL")
geneGroup <- data.frame(Group = geneGroups)
rownames(geneGroup) <- geneList

# subset vst data to include only those genes
vsd_geneList_cohort1 <- vsd_cohort1 %>%
  filter(rownames(vsd_cohort1) %in% geneList)

vsd_geneList_cohort2 <- vsd_cohort2 %>%
  filter(rownames(vsd_cohort2) %in% geneList)
```

### Euclidean distance
```{r IqbalGeneList-Euclidean-Cohort1, fig.width=14}
# Specify colors (optional). Full list of color options here: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
ann_colors_cohort1 = list(
  phenotype = c("CD8_PTCL" = "deepskyblue", "DN_PTCL" = "magenta2", "CD8_CTRL" = "seagreen3", "CD4_PTCL" = "darkkhaki", "CD4_CTRL" = "coral"),
  Group = c("TBX21-PTCL" = "steelblue", "GATA3-PTCL" = "red4"))

ann_colors_cohort2 = list(
  phenotype = c("CD4_LN_CTRL" = "coral", "CD4_PTCL" = "darkkhaki", "CD4THYM_CTRL" = "deepskyblue"),
  Group = c("TBX21-PTCL" = "steelblue", "GATA3-PTCL" = "red4"))

# Generate heatmap
vst_heatmap1 <- pheatmap(vsd_geneList_cohort1,
                        scale="row", 
                        color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
                        cluster_rows=TRUE, 
                        cluster_cols=TRUE,
                        cutree_rows = 2,
                        cutree_cols = 3,
                        clustering_distance_rows = "euclidean",
                        clustering_distance_cols = "euclidean",
                        clustering_method = "ward.D2",
                        main = "Input: Vst transformed normalized DESeq2 counts, Cohort 1\n Clustering: Ward, Distance: Euclidean",
                        annotation_col = metadata1,
                       annotation_row = geneGroup,
                       annotation_colors = ann_colors_cohort1,
                       show_rownames = TRUE) # show gene names on the heatmap
```

```{r IqbalGeneList-Euclidean-Cohort2, fig.width=16}
vst_heatmap2 <- pheatmap(vsd_geneList_cohort2,
                        scale="row", 
                        color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
                        cluster_rows=TRUE, 
                        cluster_cols=TRUE,
                        clustering_distance_rows = "euclidean",
                        clustering_distance_cols = "euclidean",
                        cutree_rows = 2,
                        cutree_cols = 3,
                        clustering_method = "ward.D2",
                        main = "Input: Vst transformed normalized DESeq2 counts, Cohort 2\n Clustering: Ward, Distance: Euclidean",
                        annotation_col = metadata2,
                       annotation_row = geneGroup,
                       annotation_colors = ann_colors_cohort2,
                       show_rownames = TRUE) # show gene names on the heatmap
```

### Pearson correlation
```{r IqbalGeneList-Pearson-Cohort1, fig.width=14}
vst_heatmap1 <- pheatmap(vsd_geneList_cohort1,
                        scale="row", 
                        color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
                        cluster_rows=TRUE, 
                        cluster_cols=TRUE,
                        clustering_distance_rows = "correlation",
                        clustering_distance_cols = "correlation",
                        cutree_rows = 2,
                        cutree_cols = 2,
                        clustering_method = "ward.D2",
                        main = "Input: Vst transformed normalized DESeq2 counts, Cohort1\nClustering: Ward, Distance: Pearson correlation",
                        annotation_col = metadata1,
                       annotation_row = geneGroup,
                       annotation_colors = ann_colors_cohort1,
                       show_rownames = TRUE) # show gene names on the heatmap
```

```{r IqbalGeneList-Pearson-Cohort2, fig.width=16}
vst_heatmap2 <- pheatmap(vsd_geneList_cohort2,
                        scale="row", 
                        color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
                        cluster_rows=TRUE, 
                        cluster_cols=TRUE,
                        clustering_distance_rows = "correlation",
                        clustering_distance_cols = "correlation",
                        cutree_rows = 2,
                        cutree_cols = 3,
                        clustering_method = "ward.D2",
                        main = "Input: Vst transformed normalized DESeq2 counts, Cohort2\nClustering: Ward, Distance: Pearson correlation",
                        annotation_col = metadata2,
                       annotation_row = geneGroup,
                       annotation_colors = ann_colors_cohort2,
                       show_rownames = TRUE) # show gene names on the heatmap
```

## Herek et al.
Source: Fig. 1A of https://doi.org/10.1182/blood.2021015019
```{r geneList-Herek}
# Define list of genes and their groups
geneList <- c("TBX21", "IFNG", "STAT1", "CSF2",
              "GATA3", "EGR1", "SEPTIN6", "CAT")
geneGroups <- c("TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL", "TBX21-PTCL",
                "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL", "GATA3-PTCL")
geneGroup <- data.frame(Group = geneGroups)
rownames(geneGroup) <- geneList

# subset vst data to include only those genes
vsd_geneList_cohort1 <- vsd_cohort1 %>%
  filter(rownames(vsd_cohort1) %in% geneList)

vsd_geneList_cohort2 <- vsd_cohort2 %>%
  filter(rownames(vsd_cohort2) %in% geneList)
```

## Euclidean distance
```{r HerekGeneList-Euclidean-Cohort1, fig.width=14}
# Specify colors (optional). Full list of color options here: http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf
ann_colors_cohort1 = list(
  phenotype = c("CD8_PTCL" = "deepskyblue", "DN_PTCL" = "magenta2", "CD8_CTRL" = "seagreen3", "CD4_PTCL" = "darkkhaki", "CD4_CTRL" = "coral"),
  Group = c("TBX21-PTCL" = "steelblue", "GATA3-PTCL" = "red4"))

# Generate heatmap
vst_heatmap1 <- pheatmap(vsd_geneList_cohort1,
                        scale="row", 
                        color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
                        cluster_rows=TRUE, 
                        cluster_cols=TRUE,
                        cutree_rows = 2,
                        cutree_cols = 2,
                        clustering_distance_rows = "euclidean",
                        clustering_distance_cols = "euclidean",
                        clustering_method = "ward.D2",
                        main = "Input: Vst transformed normalized DESeq2 counts, Cohort 1\n Clustering: Ward, Distance: Euclidean",
                        annotation_col = metadata1,
                       annotation_row = geneGroup,
                       annotation_colors = ann_colors_cohort1,
                       show_rownames = TRUE) # show gene names on the heatmap
```

```{r HerekGeneList-Euclidean-Cohort2, fig.width=16}
ann_colors_cohort2 = list(
  phenotype = c("CD4_LN_CTRL" = "coral", "CD4_PTCL" = "darkkhaki", "CD4THYM_CTRL" = "deepskyblue"),
  Group = c("TBX21-PTCL" = "steelblue", "GATA3-PTCL" = "red4"))

# Generate heatmap
vst_heatmap2 <- pheatmap(vsd_geneList_cohort2,
                        scale="row", 
                        color = colorRampPalette(c("blue", "white", "red"), space = "Lab")(100),
                        cluster_rows=TRUE, 
                        cluster_cols=TRUE,
                        cutree_rows = 2,
                        cutree_cols = 2,
                        clustering_distance_rows = "euclidean",
                        clustering_distance_cols = "euclidean",
                        clustering_method = "ward.D2",
                        main = "Input: Vst transformed normalized DESeq2 counts, Cohort 2\n Clustering: Ward, Distance: Euclidean",
                        annotation_col = metadata2,
                       annotation_row = geneGroup,
                       annotation_colors = ann_colors_cohort2,
                       show_rownames = TRUE) # show gene names on the heatmap
```

# Citations
```{r}
sessionInfo()
citation()
```


